---
title: "Sequence Analysis"
author: "Niall Newsham"
date: "31/03/2021"
output: html_document
---

Load required packages
```{r}
library(TraMineR) #for sequence analysis
library(cluster) #for cluster analysis
library(WeightedCluster) #for cluster analysis
library(tidyverse) #for data formatting
```

Read in the file containing land use data
```{r}
clusters12 <- read.csv("C:/Users/nnewsh/Documents/PhD/Projects/APPG_LBA/lsoa_12clusters.eng.csv")
```

Change land use cluster codes to labels
```{r}
clusters12[clusters12=="0"]<-"Neighbourhoods with parks"
clusters12[clusters12=="1"]<-"Neighbourhoods near countryside"
clusters12[clusters12=="2"]<-"Mixed countryside"
clusters12[clusters12=="3"]<-"Agricultural land"
clusters12[clusters12=="4"]<-"Dense central neighbourhoods"
clusters12[clusters12=="5"]<-"Noncentral neighbourhoods"
clusters12[clusters12=="6"]<-"Industrial and commercial neighbourhoods"
clusters12[clusters12=="7"]<-"Rural neighbourhoods"
clusters12[clusters12=="8"]<-"Noncentral mixed neighbourhoods"
clusters12[clusters12=="9"]<-"Pastures"
clusters12[clusters12=="10"]<-"Farmlands"
clusters12[clusters12=="11"]<-"Neighbourhoods nearby golf & other leisure"
```

tidyverse formatting. Create id column.
```{r}
clusters12 <- clusters12 %>% 
  select(-c(X)) %>% 
  mutate(id = row_number()) %>% 
  relocate(id)
```

Create a new column showing the sequences, this will be used later in the sequence analysis.
```{r}
clusters12$sequence <- paste (clusters12$clusters_12_2000, clusters12$clusters_12_2006, clusters12$clusters_12_2012, clusters12$clusters_12_2018, sep="-")
```

We are interested in LSOAs that change land use, so we will work with those only. First we create a binary variable indicating whether sequences change or not
```{r}
clusters12 %>% 
  mutate(change = case_when(
    clusters_12_2000 != clusters_12_2006 ~ "1",
    clusters_12_2000 != clusters_12_2012 ~ "1",
    clusters_12_2000 != clusters_12_2018 ~ "1")
  ) -> clusters12
```

we then filter out those that have not changed
```{r}
clusters12_change <- clusters12 %>% 
  filter(change == 1)
```

## Sequence Analysis 

First we define the sequence alphabet.
```{r}
alph <- c("Dense central neighbourhoods","Noncentral neighbourhoods",
               "Noncentral mixed neighbourhoods","Neighbourhoods with parks",
                "Rural neighbourhoods","Neighbourhoods near countryside",
               "Industrial and commercial neighbourhoods","Neighbourhoods nearby golf & other leisure", 
                "Mixed countryside","Agricultural land","Pastures","Farmlands")


scode <- c("DCN","NcN","NcMN","NwP","RUR","NnC", "IND","NnL","MC","AG","P", "FARM")
```

We then create the sequence object, specifying the values
```{r}
seq.c12 <- seqdef(clusters12_change$sequence, alphabet = alph, states = scode, cnames = c("2000", "2006", "2012", "2018"))
```

Next, we calculate the unique distance matrix using Dynamic Hamming Distances (DHD)
```{r}
seq.c12.DHD <- seqdist(seq.c12,
                       method = "DHD")
```

The next step involves clustering LSOAs based on their sequence similarity. First, we empirically test the performance of various cluster solutions.
```{r}
# PAM clustering
for (i in 6:15) {
  pamwardclust <- wcKMedoids(seq.c12.DHD, k= i)
  print(paste("number of k =", i,": Average Silhouette Width (weighted) is" , pamwardclust$stats[5]))
}
```
from this we will explore the solutions k = 11-12

PAM clustering
```{r}
pamwardclust11 <- wcKMedoids(seq.c12.DHD, k = 11)
pamwardclust12 <- wcKMedoids(seq.c12.DHD, k = 12)
```

Calculate silouette scores to use later in graphs
```{r}
sil.k11 <- wcSilhouetteObs(seq.c12.DHD, pamwardclust11$clustering, measure="ASWw")
sil.k12 <- wcSilhouetteObs(seq.c12.DHD, pamwardclust12$clustering, measure="ASWw")
```

## Plots

Before visualising the cluster solutions, we define the labels for the plots

```{r}
labs <-  c("Dense central neighbourhoods","Noncentral neighbourhoods",
           "Noncentral mixed neighbourhoods","Neighbourhoods with parks",
           "Rural neighbourhoods","Neighbourhoods near countryside",
           "Industrial and commercial neighbourhoods","Neighbourhoods nearby golf & other leisure", 
           "Mixed countryside","Agricultural land","Pastures","Farmlands")
```

Create sequence index plots

```{r, eval=FALSE}
#not run, figure is too large
seqIplot(seq.c12, 
         group = pamwardclust11$clustering,
         sortv= sil.k11,
         ltext = labs,
         cex.legend = 1,
         cex.axis=1,
         border =NA)

seqIplot(seq.c12, 
         group = pamwardclust12$clustering,
         sortv= sil.k12,
         ltext = labs,
         cex.legend = 1,
         cex.axis=1,
         border =NA)
dev.off()
```

We decide that K=11 is the most appropriate number of clusters. To save the Sequence Index plots:

```{r}
mypath <- file.path("C:/Users/nnewsh/Documents/PhD/Projects/APPG_LBA/change only lsoa/DHD/12clusters/",
                    paste("seqiplot.c12.k11.jpg", sep = ""))

jpeg(file=mypath, width = 1500, height = 1500)
seqIplot(seq.c12, 
         group = pamwardclust11$clustering,
         sortv= sil.k11,
         ltext = labs,
         cex.legend = 1,
         cex.axis=1,
         border =NA)
dev.off()
```


## TRANSITION RATES

To extract transition rates for all LSOAs:
```{r}
#Create sequence object for all LSOAs (including those that change)
seq.c12.all <- seqdef(clusters12$sequence, alphabet = alph, states = scode, cnames = c("2000", "2006", "2012", "2018"))

#examine transition rates
tr_rates.all <- seqtrate(seq.c12.all, time.varying = TRUE)
tr_rates.all <- round(tr_rates.all,3)

#save them
write.csv(tr_rates.all, "C:/Users/nnewsh/Documents/PhD/Projects/APPG_LBA/change only lsoa/DHD/12clusters/trates.all.eng.csv")

```

Transition rates for Left Behind LSOAs:

```{r}
#read in left behind areas file
LB <- read.csv("C:/Users/nnewsh/Documents/PhD/Projects/APPG_LBA/Left Behind/leftbehindLSOA.eng.csv")

#recode N/As to 0
LB <- LB %>% 
  mutate_all(~replace(., is.na(.), 0))

#record Left.Behind column to factor
LB$Left.Behind <- as.factor(LB$Left.Behind)

#merge with Clusters12 data frame
LB.merge.c12 <- merge(x = LB,
                      y = clusters12,
                      by.x = "LSOA11CD",
                      by.y = "LSOA11CD",
                      all.x = TRUE)

#we are interested in those sequences that are left behind, so we will work with those only. 
LB.merge.c12 <- LB.merge.c12 %>% 
  filter(Left.Behind == 1)

#create sequence object
seq.c12.LB <- seqdef(LB.merge.c12$sequence, alphabet = alph, states = scode, cnames = c("2000", "2006", "2012", "2018"))

#examine transition rates
tr_rates <- seqtrate(seq.c12.LB, time.varying = TRUE)
tr_rates <- round(tr_rates,3)
#save them
write.csv(tr_rates, "C:/Users/nnewsh/Documents/PhD/Projects/APPG_LBA/change only lsoa/DHD/12clusters/lb.trates.eng.csv")
```

Transition rates for non-left behind LSOAs
```{r}
#merge with Clusters12 data frame
NLB.merge.c12 <- merge(x = LB,
                      y = clusters12,
                      by.x = "LSOA11CD",
                      by.y = "LSOA11CD",
                      all.x = TRUE)

#filter out left behind LSOAs
NLB.merge.c12 <- NLB.merge.c12 %>% 
  filter(Left.Behind != 1)

#create sequence object
seq.c12.NLB <- seqdef(NLB.merge.c12$sequence, alphabet = alph, states = scode, cnames = c("2000", "2006", "2012", "2018"))

#examine transition rates
tr_rates <- seqtrate(seq.c12.NLB, time.varying = TRUE)
tr_rates <- round(tr_rates,3)
#save them
write.csv(tr_rates, "C:/Users/nnewsh/Documents/PhD/Projects/APPG_LBA/Left Behind/NLB_TR_rates_c12.csv")
```